---
title: "Case Study: {{{project_title}}}"
author:
  - name: Martin Morgan
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: Martin.Morgan@RoswellPark.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "{{{doc_date}}}"
package: "{{{pkg_ver}}}"
vignette: >
  %\VignetteIndexEntry{Case Study: {{{project_title}}} }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r internal-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
)
```

# Setup

Load useful packages

```{r setup, message = FALSE}
library(hcaCaseStudies)
library(hca)
library(dplyr)
```

## Project information

Set the project id

```{r project_id}
project_id <- "{{{project_id}}}"
```

Display information about the project

```{r project-information}
project_information(project_id)
```

## Conventions

Statistics calculated locally use `local.*` for column names, e.g.,
`local.cellCount`.

# Data processing

## Project & samples

Create a filter to identify the project

```{r filter}
project_filter <- filters(projectId = list(is = project_id))
```

Find the project

```{r find_project}
p <- projects(project_filter)
p |> glimpse()
```

Samples associated with the project

```{r samples}
s <- samples(project_filter, as = "tibble_expanded")
```

Columns are either variable and therefore describe sample-level
information...

```{r samples-exploration}
## informative 'specimen' or 'donorOrganisms' columns
s |>
    select(starts_with("specimens") | starts_with("donorOrganisms")) |>
    select(where(column_is_informative)) |>
    glimpse()
```

...or invariant and reflect experiment-wide annotation

```{r experiment-metadata}
s |>
    select(where(column_is_experimentwide)) |>
    distinct() |>
    glimpse()
```

## Data files

What types of files does the project have?

```{r files-available}
files1 <- files(project_filter)

files1 |>
    glimpse()

files1 |>
    count(fileFormat)
```

Retrieve and cache locally all files needed to construct a gene x cell
count matrix.

```{r retrieve}
file_path <-
    files1 |>
    ## filter as needed, then...
    files_download() |>
    unname()

## join file path to appropriate components of `f`
count_matrix_files1 <-
    files1 |>
    left_join(tibble(
        fileId = files1$fileId,
        local.filePath = file_path
    ), by = "fileId")
```

We can extract some data from the files, e.g., cell and gene counts

```{r file_data}
## 'files' is the final, locally modified, version of file data
count_matrix_files <-
    count_matrix_files |>
    geo_cell_count() |>
    geo_gene_count()

count_matrix_files |>
    select(name, local.cellCount, local.geneCount)
```

## Count matrix

The GSE count matrices are not quite useful, because we do not know
definitively how to map from cell to sample. We compute the count
matrix by concatenating counts from the GSM files.

```{r count-matrix}
object_name <- "local.countMatrix"
if (!cache_exists(project_id, object_name)) {
    count_matrix_files |>
        filter(startsWith(name, "GSM")) |>
        geo_count_matrix_gsms() |>
        cache_add(project_id, object_name)
}
count_matrix <- cache_read(project_id, object_name)

class(count_matrix)[1] # sparse matrix
dim(count_matrix)      # genes x cells
format(object.size(count_matrix), units = "auto")
```

## SingleCellExperiment

Column (sample) data

```{r colData}
geo_id <-
    pull(count_matrix_files, "name") |>
    strsplit("_R") |>
    vapply(head, character(1), 1L)

colData <-
    s |>
    select(where(column_is_informative)) |>
    left_join(
        bind_cols(files, local.geoId = geo_id),
        by = c(samples.id = "local.geoId")
    )
colData |>
    DT::datatable()
```

Experiment-wide metadata

```{r metadata}
metadata <-
    s |>
    select(where(column_is_experimentwide)) |>
    distinct()

metadata |>
    glimpse()
```

Assemble into SingleCellExperiment (to come)

# Anomalies

# sessionInfo() {.unnumbered}

```{r}
sessionInfo()
```
