---
title: "Case Study: A Single-Cell Transcriptomic Atlas of Human Skin Aging"
author:
  - name: Martin Morgan
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: Martin.Morgan@RoswellPark.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "6 October 2022"
package: "hcaCaseStudies 0.0.0.9004"
vignette: >
  %\VignetteIndexEntry{Case Study: A Single-Cell Transcriptomic Atlas of Human Skin Aging }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r internal-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    cache = TRUE
)
```

# Setup

Load useful packages

```{r setup, message = FALSE}
library(hcaCaseStudies)
library(hca)
library(dplyr)
```

## Project information

Set the project id

```{r project_id}
project_id <- "923d3231-7295-4184-b3f6-c3082766a8c7"
```

Display information about the project

```{r project-information}
project_information(project_id)
```

## Conventions

Statistics calculated locally use `local.*` for column names, e.g.,
`local.cellCount`.

# Data processing

## Project & samples

Create a filter to identify the project

```{r filter}
project_filter <- filters(projectId = list(is = project_id))
```

Find the project

```{r find_project}
p <- projects(project_filter, as = "tibble_expanded")
p |> glimpse()
```

Samples associated with the project

```{r samples}
s <- samples(project_filter, as = "tibble_expanded")
```

Columns are either variable and therefore describe sample-level
information... There are nine samples, ranging from 18 to 76 years
old.

```{r samples-exploration}
## informative 'specimen' or 'donorOrganisms' columns
s |>
    select(starts_with("specimens") | starts_with("donorOrganisms")) |>
    select(where(column_is_informative))
```

...or invariant and reflect experiment-wide annotation

```{r experiment-metadata}
s |>
    select(where(column_is_experimentwide)) |>
    distinct() |>
    glimpse()
```

## Data files

What types of files does the project have?

```{r files-available}
files1 <- files(project_filter)

files1 |>
    glimpse()

files1 |>
    count(fileFormat)
```

There are 72 fastq files, but no summaries, so count matrices cannot
be easily constructed.

## Row, sample, and cell annotations

We don't have any information on row (gene) data.

```{r row-data}
row_data <- tibble()
```

Column (sample & cell) data

```{r sample-data}
sample_data <-
    ## from the hca sample annotation...
    s |>
    select(
        where(column_is_informative) & where(is.atomic)
    )
    ## ...but no local computations, e.g.,gene and cell counts

sample_data
```

We don't have any cell data.

## Experiment-wide metadata

```{r metadata}
sample_metadata <-
    s |>
    select(where(column_is_experimentwide) & !any_of(names(p))) |>
    distinct()

metadata <- bind_cols(p, sample_metadata)

metadata |>
    glimpse()
```

## Assemble into SingleCellExperiment (to come)

# Anomalies

No count summary files.

# sessionInfo() {.unnumbered}

```{r}
sessionInfo()
```
