---
title: "Case Study: Robust temporal map of human in vitro myelopoiesis using single-cell genomics"
author:
  - name: Kayla Interdonanto
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: Kayla.Morrell@RoswellPark.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "1 November 2022"
package: "hcaCaseStudies 0.0.0.9005"
vignette: >
  %\VignetteIndexEntry{Case Study: Robust temporal map of human in vitro myelopoiesis using single-cell genomics }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r internal-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
)
```

# Setup

Load useful packages

```{r setup, message = FALSE}
library(hcaCaseStudies)
library(hca)
library(dplyr)
library(Seurat)
```

## Project information

Set the project id

```{r project_id}
project_id <- "ac289b77-fb12-4a6b-ad43-c0721c698e70"
```

Display information about the project

```{r project-information}
project_information(project_id)
```

## Conventions

Statistics calculated locally use `local.*` for column names, e.g.,
`local.cellCount`.

# Data processing

## Project & samples

Create a filter to identify the project

```{r filter}
project_filter <- filters(projectId = list(is = project_id))
```

Find the project

```{r find_project}
p <- projects(project_filter, as = "tibble_expanded")
p |> glimpse()
```

Samples associated with the project

```{r samples}
s <- samples(project_filter, as = "tibble_expanded")
```

Columns are either variable and therefore describe sample-level
information...

```{r samples-exploration}
## informative 'specimen' or 'donorOrganisms' columns
s |>
    select(starts_with("specimens") | starts_with("donorOrganisms")) |>
    select(where(column_is_informative)) |>
    glimpse()
```

...or invariant and reflect experiment-wide annotation

```{r experiment-metadata}
s |>
    select(where(column_is_experimentwide)) |>
    distinct() |>
    glimpse()
```

## Data files

What types of files does the project have?

```{r files-available}
files1 <- files(project_filter)

files1 |>
    glimpse()

files1 |>
    count(fileFormat)

## create `files2` with just the files needed for count-matrix
## analysis

files2 <- files1 |>
    filter(fileFormat == "rds")
```

Retrieve and cache locally all files needed to construct a gene x cell
count matrix.

```{r retrieve}
file_path <-
    files2 |>
    files_download() |>
    setNames(files2 |> pull(name))

## join file path to appropriate components of `f`
count_matrix_files1 <-
    files2 |>
    left_join(tibble(
        fileId = files2$fileId,
        local.filePath = unname(file_path)
    ), by = "fileId")
```

## Assemble into SingleCellExperiment (to come)

We can convert the RDS files (that are Seurat objects) into SingleCellExperiment
object.

```{r sce}
rds_file_paths <- count_matrix_files1 |>
    pull(local.filePath)
dat <- readRDS(rds_file_paths[2])
sce_obj <- as.SingleCellExperiment(dat)
sce_obj
```

# Anomalies

# sessionInfo() {.unnumbered}

```{r}
sessionInfo()
```
