---
title: "Case Study: A human breast atlas integrating single-cell proteomics and transcriptomics"
author:
  - name: Kayla Interdonanto
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: Kayla.Morrell@RoswellPark.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "31 October 2022"
package: "hcaCaseStudies 0.0.0.9005"
vignette: >
  %\VignetteIndexEntry{Case Study: A human breast atlas integrating single-cell proteomics and transcriptomics }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r internal-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
)
```

# Setup

Load useful packages

```{r setup, message = FALSE}
library(hcaCaseStudies)
library(hca)
library(dplyr)
```

## Project information

Set the project id

```{r project_id}
project_id <- "9b876d31-0739-4e96-9846-f76e6a427279"
```

Display information about the project

```{r project-information}
project_information(project_id)
```

## Conventions

Statistics calculated locally use `local.*` for column names, e.g.,
`local.cellCount`.

# Data processing

## Project & samples

Create a filter to identify the project

```{r filter}
project_filter <- filters(projectId = list(is = project_id))
```

Find the project

```{r find_project}
p <- projects(project_filter, as = "tibble_expanded")
p |> glimpse()
```

Samples associated with the project

```{r samples}
s <- samples(project_filter, as = "tibble_expanded")
```

Columns are either variable and therefore describe sample-level
information...

```{r samples-exploration}
## informative 'specimen' or 'donorOrganisms' columns
s |>
    select(starts_with("specimens") | starts_with("donorOrganisms")) |>
    select(where(column_is_informative)) |>
    glimpse()
```

...or invariant and reflect experiment-wide annotation

```{r experiment-metadata}
s |>
    select(where(column_is_experimentwide)) |>
    distinct() |>
    glimpse()
```

## Data files

What types of files does the project have?

```{r files-available}
files1 <- files(project_filter)

files1 |>
    glimpse()

files1 |>
    count(fileFormat)

## create `files2` with just the files needed for count-matrix
## analysis

files2 <- files1 |>
    filter(fileFormat == "csv.gz")
```

Retrieve and cache locally all files needed to construct a gene x cell
count matrix.

```{r retrieve}
file_path <-
    files2 |>
    files_download() |>
    setNames(files2 |> pull(name))

## join file path to appropriate components of `f`
count_matrix_files1 <-
    files2 |>
    left_join(tibble(
        fileId = files2$fileId,
        local.filePath = unname(file_path)
    ), by = "fileId")
```

We can extract some data from the files, e.g., cell and gene counts

```{r file_data}
## 'files' is the final, locally modified, version of file data
count_matrix_files <-
    count_matrix_files1 |>
    filter(name == "GSE180878_Li_Brugge_10XscRNAseq_GeneCellMatrix_RNAcounts_human.csv.gz") |>
    cell_count_csv() |>
    gene_count()

count_matrix_files |>
    select(name, local.geneCount, local.cellCount)
```

## Count matrix

Create a `dgCMatrix` sparse matrix (defined in the `Matrix`
package). Save it in the package cache for re-use across _R_ sessions.

<!--
add `eval = !nzchar(Sys.getenv("GITHUB_ACTIONS"))` if requires
more than 7 GB
-->

# Anomalies

No contributor generated matrices and analysis files for this project.

Ran into a 'VROOM_CONNECTION_SIZE' error. I increaded the size and reran code
only to have R crash. I believe the file is too large to process on my computer.

# sessionInfo() {.unnumbered}

```{r}
sessionInfo()
```
