---
title: "Case Study: Highly Parallel Genome-wide Expression Profiling of Individual Cells Using Nanoliter Droplets"
author:
  - name: Martin Morgan
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: Martin.Morgan@RoswellPark.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "22 September 2022"
package: "hcaCaseStudies 0.0.0.9000"
vignette: >
  %\VignetteIndexEntry{Case Study: Highly Parallel Genome-wide Expression Profiling of Individual Cells Using Nanoliter Droplets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r internal-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
)
```

# Setup

Load useful packages

```{r setup, message = FALSE}
library(hcaCaseStudies)
library(hca)
library(dplyr)
```

## Project information

Set the project id

```{r project_id}
project_id <- "3c9d586e-bd26-4b46-8690-3faaa18ccf38"
```

Display information about the project

```{r project-information}
project_information(project_id)
```

## Conventions

Statistics calculated locally use `local.*` for column names, e.g.,
`local.cellCount`.

# Data processing

## Project & samples

Create a filter to identify the project

```{r filter}
project_filter <- filters(projectId = list(is = project_id))
```

Find the project

```{r find_project}
p <- projects(project_filter)
p |> glimpse()
p |> pull(projectTitle) |> noquote()
```

Samples associated with the project

```{r samples}
s <- samples(project_filter, as = "tibble_expanded")

## all informative columns
s |>
    select(where(column_is_not_all_NA)) |>
    DT::datatable()
```

Some preliminary exploration of sample descriptions

```{r samples-exploration}
## informative 'specimen' or donorOrganisms columns
s |>
    select(starts_with("specimens") | starts_with("donorOrganisms")) |>
    select(where(column_is_informative))
```

Invariant information is really telling us about the
experiment...

```{r experiment-metadata}
s |>
    select(where(column_is_experimentwide)) |>
    distinct() |>
    glimpse()
```

## Data files

The project has 'fastq' and 'txt.gz' files. There are nine txt.gz
files but only 7 samples -- there are 7 'GSM' sample files, and two
experiment-wide 'GSE' summary files

```{r files-available}
files0 <- files(project_filter)
files0

files0 |>
    count(fileFormat)

files1 <-
    files0 |>
    filter(fileFormat == "txt.gz")
files1 |>
    select(name)
```

Retrieve and cache locally all txt.gz files

```{r retrieve}
file_path <-
    files1 |>
    filter(fileFormat == "txt.gz") |>
    files_download() |>
    unname()

files2 <-
    files1 |>
    left_join(tibble(
        fileId = files1$fileId,
        local.filePath = file_path
    ), by = "fileId")
```

We can extract some data from the files, e.g., cell and gene counts

```{r file_data}
files3 <-
    files2 |>
    cell_count_tsv() |>
    geo_gsm_gene_count()

files3 |>
    select(name, local.cellCount, local.geneCount)
```

## Count matrix

The GSE count matrices are not quite useful, because we do not know
definitively how to map from cell to sample. We compute the count
matrix by concatenating counts from the GSM files.

```{r count-matrix}
object_name <- "local.countMatrix"
if (!cache_exists(project_id, object_name)) {
    files3 |>
        filter(startsWith(name, "GSM")) |>
        count_matrix_tsv() |>
        cache_add(project_id, object_name)
}
count_matrix <- cache_read(project_id, object_name)

class(count_matrix)[1] # sparse matrix
dim(count_matrix)      # genes x cells
format(object.size(count_matrix), units = "auto")
```

## SingleCellExperiment

Column (sample) data

```{r colData}
geo_id <-
    pull(files3, "name") |>
    strsplit("_") |>
    vapply(head, character(1), 1L)
files4 <- bind_cols(files3, local.geoId = geo_id)

colData <-
    s |>
    select(where(column_is_informative)) |>
    left_join(
        bind_cols(files3, local.geoId = geo_id),
        by = c(samples.id = "local.geoId")
    )
colData |>
    DT::datatable()
```

Experiment-wide metadata

```{r metadata}
metadata <-
    s |>
    select(where(column_is_experimentwide)) |>
    distinct()

metadata |>
    t() |>
    DT::datatable()
```

Assemble into SingleCellExperiment (to come)

# Anomalies

donorOrganism.donorCount seems to be incorrect for GSM1626793 (7,
instead of 1).

```{r donorOrganisms.donorCount}
file_filter <- filters(
    projectId = list(is = project_id),
    fileFormat = list(is = "txt.gz")
)
files(file_filter, as = "tibble_expanded") |>
    select(name, donorOrganisms.donorCount, files.size)
```

# sessionInfo() {.unnumbered}

```{r}
sessionInfo()
```
