---
title: "Case Study: Single-cell Transcriptome Atlas of the Human Corpus Cavernosum"
author:
  - name: Martin Morgan
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: Martin.Morgan@RoswellPark.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "23 September 2022"
package: "hcaCaseStudies 0.0.0.9000"
vignette: >
  %\VignetteIndexEntry{Case Study: Single-cell Transcriptome Atlas of the Human Corpus Cavernosum }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r internal-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    cache = TRUE
)
```

# Setup

Load useful packages

```{r setup, message = FALSE}
library(hcaCaseStudies)
library(hca)
library(dplyr)
```

## Project information

Set the project id

```{r project_id}
project_id <- "5b910a43-7fb5-4ea7-b9d6-43dbd1bf2776"
```

Display information about the project

```{r project-information}
project_information(project_id)
```

## Conventions

Statistics calculated locally use `local.*` for column names, e.g.,
`local.cellCount`.

# Data processing

## Project & samples

Create a filter to identify the project

```{r filter}
project_filter <- filters(projectId = list(is = project_id))
```

Find the project

```{r find_project}
p <- projects(project_filter)
p |> glimpse()
```

Samples associated with the project

```{r samples}
s <- samples(project_filter, as = "tibble_expanded")
```

Columns are either variable and therefore describe sample-level
information...

```{r samples-exploration}
## informative 'specimen' or 'donorOrganisms' columns
s |>
    select(starts_with("specimens") | starts_with("donorOrganisms")) |>
    select(where(column_is_informative)) |>
    glimpse()

s |>
    count(specimens.id, specimens.disease)
```

...or invariant and reflect experiment-wide annotation

```{r experiment-metadata}
s |>
    select(where(column_is_experimentwide)) |>
    distinct() |>
    glimpse()
```

## Data files

What types of files does the project have?

```{r files-available}
files1 <- files(project_filter)

files1

files1 |>
    count(fileFormat)

files1 <-
    files1 |>
    filter(fileFormat == "tar")

files1 |>
    select(fileId, name, fileFormat)
```

Retrieve, list the content, and untar (to a temporary location; untar
is relatively quick) the file.

```{r retrieve}
tar_file <-
    files1 |>
    filter(startsWith(name, "GSE")) |>
    files_download() |>
    unname()

## what files are in the tar?
tar_files <- untar(tar_file, list = TRUE)
tar_files

## untar to a temporary location
exdir <- tempfile(); dir.create(exdir)
untar(tar_file, exdir = exdir)
```

Annotate the `files1` object with local file information

```{r file-annotate}
files_local <- tibble(
    fileId = files1$fileId,
    local.fileName = tar_files,
    local.filePath = file.path(exdir, tar_files)
)

files2 <-
    files1 |>
    ## 1:many mapping from files1 to files_local
    right_join(files_local, by = "fileId") |>
    ## make fileId unique
    mutate(fileId = paste(fileId, local.fileName, sep = "-"))
```

We can extract some data from the files, e.g., cell and gene counts

```{r file_data}
count_matrix_files <-
    files2 |>
    cell_count_csv() |>
    gene_count()

count_matrix_files |>
    select(local.fileName, local.cellCount, local.geneCount)
```

## Count matrix

Compute the count matrix by concatenating counts from the GSM files.

```{r count-matrix, eval = !nzchar(Sys.getenv("GITHUB_ACTIONS"))}
object_name <- "local.countMatrix"
if (!cache_exists(project_id, object_name)) {
    count_matrix_files |>
        count_matrix_csv() |>
        cache_add(project_id, object_name)
}
count_matrix <- cache_read(project_id, object_name)

class(count_matrix)[1] # sparse matrix
dim(count_matrix)      # genes x cells
format(object.size(count_matrix), units = "auto")
```

## SingleCellExperiment

Column (sample) data

```{r colData}
geo_id <-
    pull(count_matrix_files, "name") |>
    strsplit("_R") |>
    vapply(head, character(1), 1L)

colData <-
    s |>
    select(where(column_is_informative)) |>
    left_join(
        bind_cols(count_matrix_files, local.geoId = geo_id),
        by = c(samples.id = "local.geoId")
    )

colData |>
    DT::datatable()
```

Experiment-wide metadata

```{r metadata}
metadata <-
    s |>
    select(where(column_is_experimentwide)) |>
    distinct()

metadata |>
    glimpse()
```

Assemble into SingleCellExperiment (to come)

# Anomalies

None to report

# sessionInfo() {.unnumbered}

```{r}
sessionInfo()
```
