---
title: "Case Study: `r hcaCaseStudies::project_title('3c9d586e-bd26-4b46-8690-3faaa18ccf38')`"
author:
  - name: Martin Morgan
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: Martin.Morgan@RoswellPark.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "`r doc_date()`"
package: "`r pkg_ver('hcaCaseStudies')`"
vignette: >
  %\VignetteIndexEntry{`r hcaCaseStudies::project_title('3c9d586e-bd26-4b46-8690-3faaa18ccf38')`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r internal-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
)
```

# Setup

Load useful packages

```{r setup, message = FALSE}
library(hcaCaseStudies)
library(hca)
library(dplyr)
```

Set the project id

```{r project_id}
project_id <- "3c9d586e-bd26-4b46-8690-3faaa18ccf38"
```

Display information about the project

```{r project-information}
project_information(project_id)
```

# Data processing

## Project & samples

Create a filter to identify the project

```{r filter}
project_filter <- filters(projectId = list(is = project_id))
```

Find the project

```{r find_project}
p <- projects(project_filter)
p |> glimpse()
p |> pull(projectTitle) |> noquote()
```

Samples associated with the project

```{r samples}
s <- samples(project_filter, as = "tibble_expanded")

## all informative columns
s |>
    select(where(column_is_not_all_NA)) |>
    DT::datatable()
```

Some preliminary exploration of sample descriptions

```{r samples-exploration}
## informative 'specimen' or donorOrganisms columns
s |>
    select(starts_with("specimens") | starts_with("donorOrganisms")) |>
    select(where(column_is_informative))
```

Invariant information is really telling us about the
experiment...

```{r experiment-metadata}
s |>
    select(where(column_is_experimentwide)) |>
    distinct() |>
    glimpse()
```

## Data files

The project has 'fastq' and 'txt.gz' files. There are nine txt.gz
files but only 7 samples -- there are 7 'GSM' sample files, and two
experiment-wide 'GSE' summary files

```{r files-available}
files0 <- files(project_filter)
files0

files0 |>
    count(fileFormat)

files1 <-
    files0 |>
    filter(fileFormat == "txt.gz")
files1 |>
    select(name)
```

Retrieve and cache locally all txt.gz files

```{r retrieve}
file_path <-
    files1 |>
    filter(fileFormat == "txt.gz") |>
    files_download() |>
    unname()

files2 <-
    files1 |>
    left_join(tibble(
        fileId = files1$fileId,
        local.filePath = file_path
    ), by = "fileId")
```

We can extract some data from the files, e.g., cell and gene counts

```{r file_data}
files3 <-
    files2 |>
    geo_cell_count() |>
    geo_gene_count()

files3 |>
    select(name, local.cellCount, local.geneCount)
```
## Count matrix

```{r count-matrix}
object_name <- "local.countMatrix"
if (!cache_exists(project_id, object_name)) {
    files3 |>
        filter(startsWith(name, "GSM")) |>
        geo_count_matrix_gsms() |>
        cache_add(project_id, object_name)
}
count_matrix <- cache_read(project_id, object_name)

class(count_matrix)[1] # sparse matrix
dim(count_matrix)      # genes x cells
```

## SingleCellExperiment

Column (sample) data

```{r colData}
geo_id <-
    pull(files3, "name") |>
    strsplit("_") |>
    vapply(head, character(1), 1L)
files4 <- bind_cols(files3, local.geoId = geo_id)

colData <-
    s |>
    select(where(column_is_informative)) |>
    left_join(
        bind_cols(files3, local.geoId = geo_id),
        by = c(samples.id = "local.geoId")
    )
colData |>
    DT::datatable()
```

Experiment-wide metadata

```{r metadata}
metadata <-
    s |>
    select(where(column_is_experimentwide)) |>
    distinct()

metadata |>
    t() |>
    DT::datatable()
```

Assemble into SingleCellExperiment (to come)

# Anomalies

donorOrganism.donorCount seems to be incorrect for GSM1626793 (7,
instead of 1).

```{r donorOrganisms.donorCount}
file_filter <- filters(
    projectId = list(is = project_id),
    fileFormat = list(is = "txt.gz")
)
files(file_filter, as = "tibble_expanded") |>
    select(name, donorOrganisms.donorCount, files.size)
```

# sessionInfo() {.unnumbered}

```{r}
sessionInfo()
```
