---
title: "Case Study: The myogenesis program drives clonal selection and drug resistance in rhabdomyosarcoma"
author:
  - name: Kayla Interdonanto
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: Kayla.Morrell@RoswellPark.org
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
date: "4 November 2022"
package: "hcaCaseStudies 0.0.0.9005"
vignette: >
  %\VignetteIndexEntry{Case Study: The myogenesis program drives clonal selection and drug resistance in rhabdomyosarcoma }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r internal-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE
)
```

# Setup

Load useful packages

```{r setup, message = FALSE}
library(hcaCaseStudies)
library(hca)
library(dplyr)
library(tidyr)
```

## Project information

Set the project id

```{r project_id}
project_id <- "74e2ef9d-7c9f-418c-b281-7fb38f3b1571"
```

Display information about the project

```{r project-information}
project_information(project_id)
```

## Conventions

Statistics calculated locally use `local.*` for column names, e.g.,
`local.cellCount`.

# Data processing

## Project & samples

Create a filter to identify the project

```{r filter}
project_filter <- filters(projectId = list(is = project_id))
```

Find the project

```{r find_project}
p <- projects(project_filter, as = "tibble_expanded")
p |> glimpse()
```

Samples associated with the project

```{r samples}
s <- samples(project_filter, as = "tibble_expanded")
```

Columns are either variable and therefore describe sample-level
information...

```{r samples-exploration}
## informative 'specimen' or 'donorOrganisms' columns
s |>
    select(starts_with("specimens") | starts_with("donorOrganisms")) |>
    select(where(column_is_informative)) |>
    glimpse()
```

...or invariant and reflect experiment-wide annotation

```{r experiment-metadata}
s |>
    select(where(column_is_experimentwide)) |>
    distinct() |>
    glimpse()
```

## Data files

What types of files does the project have?

```{r files-available}
files1 <- files(project_filter)

files1 |>
    glimpse()

files1 |>
    count(fileFormat)

## create `files2` with just the files needed for count-matrix
## analysis

files2 <- files1 |>
    filter(fileFormat == "tar")
```

Retrieve and cache locally all files needed to construct a gene x cell
count matrix.

```{r retrieve}
tar_file <-
    files2 |>
    files_download() |>
    setNames(files2 |> pull(name))
```

This is a 'tar' archive of GSM files; untar these to a temporary
location

```{r}
## what files are in the tar?
tar_files <- untar(tar_file, list = TRUE)
length(tar_files)
head(tar_files)
tar_files <- tar_files[!grepl("fragments.tsv.gz.tbi", tar_files)]
## untar to a temporary location
exdir <- tempfile(); dir.create(exdir)
untar(tar_file, exdir = exdir)
```

Parse the file names to create a tibble with local file path, file
type, and sample id

```{r local-files}
local_files <-
    tibble(
        local.filePath = file.path(exdir, tar_files),
        ## local.fileType becomes one of barcodesPath, genesPath, matrixPath
        local.fileType = sub(".*_([[:alpha:]]*).*", "\\1Path", tar_files),
        local.sampleId = sub("([^_]*).*", "\\1", tar_files)
    ) |>
    ## each 'sampleId' has a barcodesPath, genesPath, and matrixPath
    pivot_wider(
        names_from = local.fileType,
        names_prefix = "local.",
        values_from = "local.filePath"
    )
## 64 samples
local_files |>
    mutate(across(ends_with("Path"), basename))
```

## Count matrix

Create a `dgCMatrix` sparse matrix (defined in the `Matrix`
package). Save it in the package cache for re-use across _R_ sessions.

<!--
add `eval = !nzchar(Sys.getenv("GITHUB_ACTIONS"))` if requires
more than 7 GB
-->


```{r file-data}
## Remove files with NA in featuresPath and/or barcodesPath
local_files2 <- local_files |>
    filter(!(is.na(local.featuresPath)))
gene_and_cell_counts <-
    local_files2 |>
    csv_tsv_count_lines("local.sampleId", "local.barcodesPath") |>
    rename(local.cellCount = local.lines) |>
    csv_tsv_count_lines("local.sampleId", "local.featuresPath") |>
    rename(local.geneCount = local.lines)
gene_and_cell_counts |>
    select(local.sampleId, local.geneCount, local.cellCount)
```

```{r count-matrix}
object_name <- "local.countMatrix"
if (!cache_exists(project_id, object_name)) {
    local_files2[1:20,] |> ## processing first 20 since there seemed to be an issue with file 21
        mtx_count_matrix(
            "local.matrixPath", "local.featuresPath", "local.barcodesPath"
        ) |>
        cache_add(project_id, object_name)
}
count_matrix <- cache_read(project_id, object_name)

class(count_matrix)[1] # sparse matrix
dim(count_matrix)      # genes x cells
format(object.size(count_matrix), units = "auto")
```

## Row, sample, and cell annotations

Row (gene) data

```{r row-data}
row_data <- 
    local_files2 |>
    slice(1) |>
    pull("local.featuresPath") |>
    readr::read_tsv(
        col_names = c("ensembl_id", "hugo_symbol", "annotation"),
        show_col_types = FALSE
    )

row_data
```

Column (sample & cell) data

```{r sample-data}
sample_data <-
    ## from the hca sample annotation...
    s |>
    select(
        where(column_is_informative) & where(is.atomic)
    ) |>
    ## ...and local computations, e.g.,gene and cell counts
    left_join(
        gene_and_cell_counts |>
        select(local.sampleId, local.cellCount, local.geneCount),
        by = c(samples.id = "local.sampleId")
    )

sample_data
```

```{r cell-data}
cell_barcodes <- 
    local_files2 |>
    pull("local.barcodesPath") |>
    lapply(readLines)

sampleId <- pull(local_files2, "local.sampleId")
cell_data <- tibble(
    local.sampleId = rep(sampleId, lengths(cell_barcodes)),
    local.barcodes = unlist(cell_barcodes, use.names = FALSE)
)

cell_data

## N.B. some barcodes are duplicated multiple times
cell_data |>
    count(local.barcodes, sort = TRUE, name = "occurrences") |>
    count(occurrences, name = "n_barcodes")
```

## Experiment-wide metadata

```{r metadata}
sample_metadata <-
    s |>
    select(where(column_is_experimentwide) & !any_of(names(p))) |>
    distinct()

metadata <- bind_cols(p, sample_metadata)

metadata |>
    glimpse()
```

## Assemble into SingleCellExperiment (to come)

# Anomalies

# sessionInfo() {.unnumbered}

```{r}
sessionInfo()
```
